Testable normative theories

### 2020-05-25 simple.tex
Normative theory separated from empirical one

\subsection{Definition}
A normative theory is a tuple $(\allprops, ¬, \allargs, I, (\ileadstoe[i])_{i \in I}, q, (\gleadsto[q(i)])_{q(i) \in q(I)}, \gbeats, +)$,
with ${\ileadstoe[i]} \subseteq (\allargs × \allargs) × \Phi$, ${\gleadsto[q(i)]} \subseteq \allargs × \Phi$, ${\gbeats} \subseteq \allargs × \allargs$,
where $\Phi = \allprops × \set{\text{possible}, \text{sure}}$, and $+$ a commutative associative internal binary operation on $\allargs$. Given $\args \subseteq \allargs$, the notation $\sum \args = \sum_{\ar \in \args} \ar$ thus designates a well-defined argument.

@@

To lighten notations, $\ileadstoe[i]$ will generally be denoted by $\ileadstoe$, and $\gleadsto[q(i)]$ by $\gleadsto$, with the understanding that these symbols result of applying functions of, respectively, $i$ and $q(i)$. When the symbol $i$, $\ileadstost$ or $\gleadsto$ is used in a formula without $i$ being bound, the binding $\forall i \in I$ is implied.

@@

Given $\phi \in {\gleadsto}(\allargs)$, define ${\ibeatse[!\phi]} \subseteq \allargs × \allargs$ such that $\ar' \ibeatse[!\phi] \ar$ iff $(\ar', \ar) \ileadstoe ¬\phi$.

@@

\commentOC{TODO next: check $\gbeatsstar$ can be replaced. Then, comment that Bounded length does not imply finiteness of arguments used as counter-arguments by the model: model is able to counter infinitely many arguments with infinitely many counter-args, but needs to declare that an (unknown) finite subset of those will suffice to convince $i$.}

It is desirable to check that \cref{ax:nocontr,ax:norm} do not hide a requirement on the content of $\ileadstostp$, thus, that they are together satisfiable without constraining $\ileadstostp$. To define this more precisely, consider temporarily these axioms as properties of a set $\iPhi \subseteq \Phi$. I write $A^1(\iPhi)$ if “axiom” 1 (interpreted as a property of $\iPhi$) holds on $\iPhi$, and similarly for $A^2(\iPhi)$. The required check amounts to prove that $\exists \iPhi \subseteq \Phi \suchthat A^1(\iPhi) \land A^2(\iPhi)$.

Given a model, define $\Box\Phi \subseteq \Phi$ as 
$\propposs \in \Box\Phi$ iff $[\exists \ar \in \allargs \suchthat \ar \ileadstostp \propposs]$
and 
$\propsure \in \Box\Phi$ iff $[\exists \ar \in \allargs \suchthat \ar \ileadstostp \propsure \land \ar \ileadstostp \propposs]$.

The propositions such that $\phi \in \Box\Phi \lor ¬\phi \in \Box\Phi$ are those on which the model permits to take a position. For that subset of propositions (which can be empty): $\phi \in \Box\Phi \oplus ¬\phi \in \Box\Phi$, where $\oplus$ denotes the exclusive disjunction; equivalently: $\phi \in \Box\Phi ⇔ ¬\phi \notin \Box\Phi$. This is proven below, and shows that \cref{ax:nocontr,ax:norm} are satisfiable together (suffices to define $\iPhi$ as the minimal set satisfying \cref{ax:norm}). Also proven below is that, for that same subset of propositions: $\phi \in \iPhi \oplus ¬\phi \in \iPhi$, which ensures an unambiguous interpretation of the content of $\iPhi$ for those propositions whose status is known thanks to the model.

\subsection{Validity and conditions}
\begin{definition}[Validity]
	A theory is \emph{valid} iff $\forall \phi \in {\gleadsto}(\allargs): [\phi \in \iPhi \land ¬\phi \notin \iPhi]$.
\end{definition}

@@

Define $\mbeats[g] \subseteq \gargs × \allargs$ as $\ar_2 \mbeats[g] \ar_1$ iff $\exists \ar \in c(\ar_2) \suchthat \ar \mbeats \ar_1$.
Define $\mbeats[a] \subseteq \allargs × \gargs$ as $\ar_1 \mbeats[a] \ar_0$ iff $\exists \ar \in c(\ar_0) \suchthat \ar_1 \mbeats \ar$ and $\nexists \ar \in c(\ar_0) \suchthat \ar \mbeats \ar_1$. Define ${\gbeatsstar} \subseteq \allargs × \allargs$ as ${\gbeatsstar} = {\mbeats[g]} \cup {\mbeats[a]}$. \commentOC{TODO $c^\phi$, not $c$. Use $c = \cup_{\phi} c^\phi$?}

@@

The following condition mandates that if the arguments used by the theory are not decisive, it knows a reason.
\begin{definition}[Answerability]
	$\forall \phi \in {\gleadsto}(\allargs), \ar_0 \in \gargs[\phi], \ar_1 \in \allargs: (\ar_1, \ar_0) \nileadstost \phi ⇒ [\exists \ar_1' \in {\gbeatsstarinv}(\ar_0)]$.
\end{definition}

@@

\begin{definition}[No repetition]
	$\nexists \ar_0 \in \gargs[\phi], \ar_1 \in \allargs, \ar_2 \in c(\ar_0) \suchthat \ar_2 \gbeats \ar_1 \gbeatsstar \ar_0$.
\end{definition}

\begin{definition}[Operational validity]
	$\forall \phi \in {\gleadsto}(\allargs), \ar_0 \in \gargs[\phi], \ar_1 \in \allargs: [\ar_1 \gbeatsstar \ar_0 \land (\ar_1, \ar_0) \nileadstost \phi] ⇒ \exists \ar_2 \in {\gbeatsinv}(\ar_1)$.
\end{definition}

\subsection{Theorem}
\begin{theorem}
	If a theory is answerable, has bounded length, does not repeat, and is operationally valid, it is valid.
\end{theorem}
\begin{proof}
	Lemma: given $\phi \in {\gleadsto}(\allargs)$ and $\ar[k] \in \gargs[\phi]$, if $\ar[k] \nileadstost \phi$, then: $\exists \ar[k + 1] \in \gargs[\phi] \suchthat \card{c(\ar[k + 1])} ≥ \card{c(\ar[k])} + 1$.
	To prove the lemma, let us pick any $\phi \in {\gleadsto}(\allargs)$ and $\ar[k] \in \gargs[\phi]$ such that $\ar[k] \nileadstost \phi$. 
	We know that $\exists \ar_1 \in \allargs \suchthat (\ar_1, \ar[k]) \nileadstost \phi$. 
	Therefore, by Answerability, $\exists \ar_1' \in \allargs \suchthat \ar_1' \gbeatsstar \ar[k]$, and, by Operational validity, $\exists \ar_2 \in \gargs[\phi] \suchthat \ar_2 \gbeats \ar'_1$. 
	Define $\ar[k + 1] = \ar[k] + \ar_2$. 
	By No repetition, $\ar_2 \notin c(\ar[k])$.
	Thus, $\card{c(\ar[k + 1])} ≥ \card{c(\ar[k])} + 1$.

@@

@@

Thus, four variants, given $\ar_2 \gbeats \ar_1$.
\begin{itemize}
	\item Strong answ with single try: $\ar_1 \notin f(\ar_0 + \ar_2) \land \forall \ar_1' \in f(\ar_0 + \ar_2): \ar_1' \gbeatsstar \ar_0 + \ar_2$.
	\item Strong answ with multiple tries: $\forall \ar_1' \in f(\ar_0 + \ar_2): \ar_1' \gbeatsstar \ar_0 + \ar_2$. (Example: $\ar_1 = \text{“for ethical reasons”}$, and $\eta$ knows two c-a to this, from different angles.) 	\item Weak answ with single try: $\ar_1 \notin f(\ar_0 + \ar_2) \land [f(\ar_0 + \ar_2) ≠ \emptyset ⇒ \exists \ar_1' \suchthat \ar_1' \gbeats \ar_0]$.
	\item Weak answ with multiple tries: $f(\ar_0 + \ar_2) ≠ \emptyset ⇒ \exists \ar_1' \suchthat \ar_1' \gbeats \ar_0 + \ar_2$.
\end{itemize}

@@

TODO false: we try any sequence and then poof. Thus, is my approach insufficiently general (and thus should be restrained to the single-try models anyway)?

\begin{definition}[No self-attack]
	$\gbeatsstar$ must be asymmetric.
\end{definition}

\begin{remark}
	When $\gbeatsstar$ is as defined above, asymmetry of $\gbeatsstar$ with the other conditions implies No repetition: by construction of $\gbeatsstar$, $\forall \ar_0 \in \gargs[\phi], \ar_1 \in \allargs, \ar_2 \in c(\ar_0) \suchthat \ar_2 \gbeats \ar_1: \ar_0 \gbeatsstar \ar_1$, thus, by asymmetry, $¬(\ar_1 \gbeatsstar \ar_0)$.
\end{remark}

>>>>

\subsection{Definition}
A normative theory is a tuple $(\allprops, ¬, \allargs, I, (\ileadstoe[i])_{i \in I})$, 
with ${\ileadstoe[i]} \subseteq (\allargs × \allargs) × \Phi$, 
where $\Phi = \allprops × \set{\text{possible}, \text{sure}}$.

@@

To lighten notations, $\ileadstoe[i]$ will generally be denoted by $\ileadstoe$, this symbol being understood as the result of a function of $i$. When the symbol $i$, $\ileadstoe$ or $\ileadstost$ is used in a formula without $i$ being bound, the binding $\forall i \in I$ is implied.

@@

For analytical clarity, it is desirable to check that \cref{ax:nocontr,ax:norm} do not hide a requirement on the content of $\ileadstostp$, thus, that they are together satisfiable without constraining $\ileadstostp$. To define this more precisely, consider temporarily these axioms as properties of a set $\iPhi \subseteq \Phi$. I write $A^1(\iPhi)$ if “axiom” 1 (interpreted as a property of $\iPhi$) holds on $\iPhi$, and similarly for $A^2(\iPhi)$. The required check amounts to prove that $\exists \iPhi \subseteq \Phi \suchthat A^1(\iPhi) \land A^2(\iPhi)$.

Given a normative theory, define $\Box\Phi \subseteq \Phi$ as 
$\propposs \in \Box\Phi$ iff $[\exists \ar \in \allargs \suchthat \ar \ileadstostp \propposs]$
and 
$\propsure \in \Box\Phi$ iff $[\exists \ar \in \allargs \suchthat \ar \ileadstostp \propsure \land \ar \ileadstostp \propposs]$.
Define  $P^* \subseteq \Phi$ as the set of all \emph{decidable} propositions: $P^* = \set{\phi \in \Phi \suchthat \phi \in \Box\Phi \lor ¬\phi \in \Box\Phi}$; and define $P \subseteq \Phi$ as decidable iff $P \subseteq P^*$. 
The propositions in $P^*$ are those on which the normative theory permits to take a position. It will generally be unknown in practical applications, and may be empty.

For any decidable proposition: $\phi \in \Box\Phi \oplus ¬\phi \in \Box\Phi$, where $\oplus$ denotes the exclusive disjunction; equivalently: $\phi \in \Box\Phi ⇔ ¬\phi \notin \Box\Phi$. This is proven below, and shows that \cref{ax:nocontr,ax:norm} are satisfiable together (suffices to define $\iPhi = \Box\Phi$). Also proven below is that, for that same subset of propositions: $\phi \in \iPhi \oplus ¬\phi \in \iPhi$, which ensures an unambiguous interpretation of the content of $\iPhi$ for those propositions whose status is known thanks to the model.

\section{Empirical theories}
\subsection{Definition}
An empirical theory for a normative theory is a tuple $(q, (\gleadsto[q(i)])_{q(i) \in q(I)}, \gbeats, +)$,
with ${\gleadsto[q(i)]} \subseteq \allargs × \Phi$, ${\gbeats} \subseteq \allargs × \allargs$,
and $+$ a commutative associative internal binary operation on $\allargs$. Given a finite $\args \subseteq \allargs$, the notation $\sum \args = \sum_{\ar \in \args} \ar$ thus designates a well-defined argument. 
\commentOC{Check about finiteness.}

An empirical theory is always defined for some normative theory. In the following, any reference to an empirical theory should always be understood as a reference to an implicit normative theory and an empirical theory for that normative theory.

As above,  $\gleadsto$ will be used instead of $\gleadsto[q(i)]$ when it creates no ambiguity, with the understanding that this symbol result of applying a function of $q(i)$; and the binding $\forall i \in I$ will generally be omitted.

Given $\phi \in {\gleadsto}(\allargs)$, define ${\ibeatse[!\phi]} \subseteq \allargs × \allargs$ such that $\ar' \ibeatse[!\phi] \ar$ iff $(\ar', \ar) \nileadstost \phi$.

@@

\commentOC{TODO also: in general, the model defines $\gargs[\phi]$ as a function of $\ileadstoe$? Start with $\gargs[\phi, 0]$, then observe all $\ar_1$ attacking those, and the model may define $\gargs[\phi, 1]$ depending on this observation, and so on. The condition is that one such argument be decisive.}

\subsection{Validity and conditions}
\begin{definition}[Validity]
	An empirical theory is \emph{valid} iff $\forall \phi \in {\gleadsto}(\allargs): [\phi \in \iPhi \land ¬\phi \notin \iPhi]$.
\end{definition}

@@

\begin{remark}
	In fact, the model is tasked with finding a stable point after interrogating $i$. The process of trying counter-arguments $\ar_1, …$ is $q(i)$. When done, the model claims that $\ar_0 + \ar_2 + …$ is decisive. The model also claims that another process of interrogation does not lead to different answers. This is falsifiable.

	It follows that trying different $\ar_2, \ar'_2, …$ is also admissible.

	I need a (more) basic observable: a path $p$ is a sequence of arguments given to $i$, and I assume that whether $p \ileadsto[i] \phi$ is observable. But we can only assume that one path and its extensions is observable.

	When claiming that $(\ar_1, \ar_0) \ileadstost[i] \phi$, the model really claims that $\nexists p, i' \in q^{-1}(q(i)) \suchthat p, \ar_1, \ar_0 \ileadsto[i'] ¬\phi \lor p, \ar_0, \ar_1 \ileadsto[i'] ¬\phi$, where $i' \in q^{-1}(q(i))$ iff $q(i) = q(i')$.

	To say that $\phi \in \iPhi$, need to not depend on choice of $q$, in the following sense: if with $q$, $\phi$ seems stable, but with $q'$, $¬\phi$ seems stable, then $\phi$ is not proved as being in $\iPhi$. But if with $q$, $\phi$ seems stable, but with $q'$, nothing can be shown to be stable, then $\phi$ is proved (temporarily) as being in $\iPhi$. More precisely, after $q'$, the model must be able to stabilize $i$ with its $q$.
\end{remark}

@@

The following condition mandates that if the arguments used by the theory are not decisive, it knows a reason.
\begin{definition}[Answerability]
	$\forall \phi \in {\gleadsto}(\allargs), \ar_0 \in \gargs[\phi], \ar_1 \in \allargs: (\ar_1, \ar_0) \nileadstost \phi ⇒ [\exists \ar_1' \in {\gbeatsinv}(\ar_0)]$.
\end{definition}

@@

\begin{remark}
	Bounded length does not imply finiteness of arguments used as counter-arguments by the model: model is able to counter infinitely many arguments with infinitely many counter-args, but needs to declare that an (unknown) finite subset of those will suffice to convince $i$.
\end{remark}

\begin{definition}[No repetition]
	$\nexists \ar_0 \in \gargs[\phi], \ar_1 \in \allargs, \ar_2 \in c(\ar_0) \suchthat \ar_2 \gbeats \ar_1 \gbeats \ar_0$.
\end{definition}

\begin{definition}[Operational validity]
	$\forall \phi \in {\gleadsto}(\allargs), \ar_0 \in \gargs[\phi], \ar_1 \in \allargs: [\ar_1 \gbeats \ar_0 \land (\ar_1, \ar_0) \nileadstost \phi] ⇒ \exists \ar_2 \in {\gbeatsinv}(\ar_1)$.
\end{definition}

I want an empirical theory to: 1) declare some $P \subseteq \Phi$ and it being valid; 2) prove its claims (otherwise, uncheckable and unfalsifiable and thus not empirical) by producing decisive arguments; 3) permit reuse… Focusing on valid theories is thus justified by the fact that if no valid theory for some $P$, then any theory will not be able to prove its claims (as there is no decisive argument).

\subsection{Theorem}
\begin{theorem}
	If a theory is answerable, has bounded length, does not repeat, and is operationally valid, it is valid.
\end{theorem}
\begin{proof}
	Lemma: given $\phi \in {\gleadsto}(\allargs)$ and $\ar[k] \in \gargs[\phi]$, if $\ar[k] \nileadstost \phi$, then: $\exists \ar[k + 1] \in \gargs[\phi] \suchthat \card{c(\ar[k + 1])} ≥ \card{c(\ar[k])} + 1$.
	To prove the lemma, let us pick any $\phi \in {\gleadsto}(\allargs)$ and $\ar[k] \in \gargs[\phi]$ such that $\ar[k] \nileadstost \phi$. 
	We know that $\exists \ar_1 \in \allargs \suchthat (\ar_1, \ar[k]) \nileadstost \phi$. 
	Therefore, by Answerability, $\exists \ar_1' \in \allargs \suchthat \ar_1' \gbeats \ar[k]$, and, by Operational validity, $\exists \ar_2 \in \gargs[\phi] \suchthat \ar_2 \gbeats \ar'_1$. 
	Define $\ar[k + 1] = \ar[k] + \ar_2$. This is the element we look for.
	First, $\ar[k + 1] \in \gargs[\phi]$ because $\ar[k + 1] \gbeats \ar_1' \ibeatse[!\phi] \ar[k]$.
	Second, by No repetition, $\ar_2 \notin c(\ar[k])$,
	thus, $\card{c(\ar[k + 1])} ≥ \card{c(\ar[k])} + 1$ (observing that, by definition of $c$, for any $\ar, \ar'$: $c(\ar) \subseteq c(\ar + \ar')$).

@@

\subsection{Multiple tries}
Should generalize the building of possible arguments attempted by the model. A model defines $m(\ar_0, \ar_1) \subseteq \allargs$ giving its possible arguments for answering $\ar_1$. Define $\gargs[\phi, 0] = \mleadsto(\allargs)$. If $\ar_k$ has lost (thus the model can’t answer some counter-argument to it), there must be another candidate argument.

Mandate that $\exists \ar_0 \in \gargs[\phi, 0]$ such that $\exists \ar_1 \in \allargs \suchthat \ar_1 \ibeatse[!\phi] \ar \land f(\ar_0, \ar_1) ≠ \emptyset$.

\commentOC{TODO 1. Justify the fundamental restriction: mandates that $\exists \ar_0 \in \gargs[\phi] \suchthat \ar_0$ decisive. 2. Rephrase condition as: must not be that at some point, for all $\ar_0$ in the current set of possibilities, $\nexists \ar_1 \ibeatse \ar_0 \suchthat \ar_2 \gbeats \ar_1$ (or rather $m(\ar_0, \ar_1) \subseteq C$ with $C$ the current set of possibilities).}

First point is not restrictive: if there is a decisive argument for $\phi$, there is a valid model which says that $\phi$ (and if there is a valid model for $\phi$, then $\phi \in \iPhi$).

First point bars models that want to fully explore $\ibeatse$ and then choose a decisive argument, for example. This would not be desirable because if model can’t be a priori expressed as a restricted model, then we have no means to check that really $\ar \ileadsto \phi$ as another path can’t be adopted (unless $i$ forgets). Also, what if $i$ gets tired after 10 arguments?

@@

Thus, four variants, given $\ar_2 \gbeats \ar_1$.
\begin{itemize}
	\item Strong answ with single try: $\ar_1 \notin f(\ar_0 + \ar_2) \land \forall \ar_1' \in f(\ar_0 + \ar_2): \ar_1' \gbeats \ar_0 + \ar_2$.
	\item Strong answ with multiple tries: $\forall \ar_1' \in f(\ar_0 + \ar_2): \ar_1' \gbeats \ar_0 + \ar_2$. (Example: $\ar_1 = \text{“for ethical reasons”}$, and $\eta$ knows two c-a to this, from different angles.) 	\item Weak answ with single try: $\ar_1 \notin f(\ar_0 + \ar_2) \land [f(\ar_0 + \ar_2) ≠ \emptyset ⇒ \exists \ar_1' \suchthat \ar_1' \gbeats \ar_0]$.
	\item Weak answ with multiple tries: $f(\ar_0 + \ar_2) ≠ \emptyset ⇒ \exists \ar_1' \suchthat \ar_1' \gbeats \ar_0 + \ar_2$.
\end{itemize}

@@

\commentOC{This is incorrect. The current approach already mandates single-try. If $\ar_2 \gbeats \ar_1$, then $\ar_1 \ibeatse \ar_0 + \ar_2$, in the sense that $\ar_1$ would be the only such argument, is forbidden because this would require $\ar_1 \gbeats \ar_0 + \ar_2$ and thus yields a repetition.}
